A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-14 tex/codatadp.tex
More reorg, will rework text.

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}
\newcommand{\vsimv}[1]{\vsim_{#1}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}

@@

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node $\Fc$,
then $\Ec$ is satisfiable $\thD$.
We demonstrate a particular model $\M$ of $\thD$ that satisfies $\Ec$.
Before doing so, we require additional terminology concerning $\mu$-terms.

For a $\mu$-term $t_1$ with subterm $t_2$, 
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec u ) \in t_1$ indicates that $\mu x. C( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
t_1\interp{x}^M & = & \mu x. C( t_1\interp{ \vec u}^{M \cup \{ x \}}) & x \not\in M, x \mapsto \mu x. C( \vec u ) \in t_1  \\
t_1\interp{x}^M & = & x & \text{otherwise} \\
t_1\interp{\mu x. C( \vec u )}^M & = & \mu x. C( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & x \not\in M \\
t_1\interp{\mu x. C( \vec u )}^M & = & x & \text{otherwise}
\end{array}\]
We will often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.
We will write $\mathcal{T}( t_1 )$ to denote the set consisting of $t_1\interp{t_2}$ for all subterms $t_2$ of $t_1$,
and write $\mathcal{T}( S )$ to denote the union of $\mathcal{T}( t )$ for each $t \in S$ where $S$ is a set of terms.

Given a $\mu$-term $t_1$ of the form $\mu x. C( u_1, \ldots, u_n )$ with subterm $t_2$,
we say $t_2$ is a \emph{bisimilar subterm} of $t_1$ if and only if it is of the form
$\mu y. C( v_1, \ldots, v_n )$ and $t_1\interp{ u_i } \sim t_1\interp{ v_i }$ for $i = 1,\ldots, n$.
We say $t_1$ is \emph{normal} if and only if it does not contain a strict bisimilar subterm
and all of its strict subterms are also normal.
For instance, the $\mu$-term $t_1$ of the form $\mu x. \const{S}( \mu y. \const{S}( y ) )$ is not normal, 
since $\mu y. \const{S}( y )$ is a bisimilar subterm of $t_1$.
In particular, note that their arguments have the same interpretation in $t_1$, 
that is,
$t_1\interp{\mu y. \const{S}( y )}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{y}^{\{y\}} ) =$ 
$\mu y. \const{S}( y )$ 
is syntactically equivalent to
$t_1\interp{y}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{y}^{\{y\}} ) =$ 
$\mu y. \const{S}( y )$.
The $\mu$-term $t_1$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is also not normal, 
since $\mu y. \const{S}( x )$ is a bisimilar subterm of $t_1$,
noting that
$t_1\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( t_1\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( t_1\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$
is syntactically equivalent to $t_1\interp{x}^\emptyset = t_1$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form} of $v$)
that is a renaming of $v$ by replacing all of the bisimilar subterms of $v$ with $x$,
and by replacing all of the non-normal subterms of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.

\begin{definition}[Normal Model] \label{def:norm-model}
A model $\M$ is \emph{normal} if and only if:
\begin{enumerate}
\item 
for all types $\tau$, 
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are 
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$,
\item 
for all constructor terms $\const{C}( \vec t )$ of type $\tau$,
$\M( \const{C}( \vec t ) )$ is the value 
in $\M( \tau )$ that is a renaming of 
$\nf{(\mu x. \const{C}( \M( \vec t ) ))}$ where $x \not\in FV( \vec t )$, and
\item 
for all selector terms $s^j_i( t )$ of type $\tau$ where $\M( t )$ is of the form $\mu x. \const{C}_i( \vec u )$.% for some $\vec u$,
$\M( s^j_i( t ) )$ is the value
in $\M( \tau )$ that is a renaming of 
$\M( t )\interp{u_j}$.
\end{enumerate}
\end{definition}

In the following, we consider only normal models.
Thus, when constructing any model $\M$,
it remains only to define how wrong-applied selector terms and the free variables are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
We first state the following property of $\Val$, which amounts to saying that interpretation of an equivalence class is
not sensisitive to the context in which it is interpretated.

\begin{lemma}
If $\Val$ is a mapping construced in step two of our calculus,
then $t_1 \interp{ \mu v_{\ec{x}}. t_2 } = \Val \ec{x}$
for all $t_1\interp{\mu v_{\ec{x}}. t_2} \in \mathcal{T}(\Val)$.
\end{lemma}
\begin{proof}
\rem{TODO}
\qed
\end{proof}

We say $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing any variable besides $x$.
For instance, for $t:= \mu x. \const{C}( x, y )$ and $u := \mu y. \const{C}( x, y )$,
we have $t \not\vsim u$, but it is the case that both
$t \vsimv{x} u$ and $t \vsimv{y} u$.

The \emph{completion} of $\Val$, which we will write as $\Val^\ast$, 
is the result of exhaustively applying the following update to $\Val$:

\(
\inferrule{
  x : \tau \in FV( \Val ) 
  \quad
  c \in \M( \tau )
  \quad
  c \not\sim_x t \text{ for all } t \in \mathcal{T}(\vrange(\Val))
}{
  \Val := \Val \{ x \mapsto c \}
}
\)

\rem{revise from here on}

\begin{lemma} \label{lem:normal-subterm-bisim}
If $t_1 \vsim t_2$, then $t_1\interp{\tpos{t_1}{p}} \vsim t_2\interp{\tpos{t_2}{p}}$.
\end{lemma}
\begin{proof}
\end{proof}

\begin{lemma} \label{lem:normal-subs}
If $t_1$ and $t_2$ are normal $\mu$-terms where $t_1 \not\vsim t_2$,
and if $u_1$ is a normal $\mu$-term such that $u_1 \not\vsim u_2 \sigma$ for all $\sigma$, $u_2 \in \mathcal{T}( t_1 ) \cup \mathcal{T}( t_2 )$,
then $\nf{(t_1 \sigma_x)} \not\vsim \nf{(t_2 \sigma_x)}$ for $\sigma_x := \{ x \mapsto u_1 \}$.
\end{lemma}
\begin{proof}
By contradiction assume that 
$t_1 \not\vsim t_2$ and 
$\nf{(t_1 \sigma)} \vsim \nf{(t_2 \sigma)}$
for some $\sigma := \{ x \mapsto u_1 \}$
where $u_1 \not\vsim u_2 \sigma$ for all $\sigma$, $u_2 \in \mathcal{T}( t_1 ) \cup \mathcal{T}( t_2 )$ 
and $t_1, t_2, u_1$ are normal.
Then it must be the case that there is a position where, say $t_1$, 
is such that $\tpos{t_1}{p} = x$ and $\tpos{t_2}{p} = \mu y. u$,
and by Lemma~\ref{lem:normal-subterm-bisim}, where $\tpos{\nf{(t_1 \sigma)}}{p} \vsim \tpos{\nf{(t_2 \sigma)}}{p}$.

%Since $\nf{(t_2 \interp{\mu y. u})} \in \mathcal{T}( t_2 )$.
\end{proof}

@@

We construct a model $\M$ of $\thD$ that interprets each type $\tau$
as a maximal set of closed normal $\mu$-terms of that type that are 
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$.

The interpretation of constructor terms
$\M( C( \vec t ) )$ for $C \in \Ctr$ is the value in $\M$ that is a renaming of $\nf{(\mu x. C( \M( \vec t ) ))}$ where $x \not\in FV( \vec t )$.
The interpretation of (correctly-applied) selector terms
$\M( s^j_i( t ) )$ for $s^j_i \in \Sel$ is the value in $\M$ that is a renaming of $\M( t )\interp{u_j}$
where $\M( t )$ is of the form $\mu x. C_i( \vec u )$ for some $\vec u$.
For instance, if $\M( t ) = \mu x. \const{C}_1( x )$ (call this term $v$), 
then $\M( s_1^1( t ) )$ is a renaming of
$v\interp{x} =$ 
$v\interp{\mu x. \const{C}_1( x )}^\emptyset =$
$\mu x. \const{C}_1( v\interp{x}^{\{x\}} ) =$
$\mu x. \const{C}_1( x )$.
%To interpret a correctly-applied selector in $\M$, if $v = \mu x. C_i( \vec t )$, then $\M( s^j_i( v ) )$ is the value in $\M( \tau )$ that is a renaming of $v\interp{t_j}$,
%where $t_j$ has type $\tau$.
%We assume $\M$ interprets constructors and correctly-applied selectors in the obvious way
%\footnote{
%The interpretation of $\M( C( \vec t ) )$ is the value in $\M( \tau )$ that is a renaming of $\nf{\mu x. C( \M( \vec t ) ) }$.
%To interpret a correctly-applied selector, if $v = \mu x. C_i( \vec t )$, then $\M( s^j_i( v ) )$ is the value in $\M( \tau )$ that is a renaming of $v\interp{t_j}$,
%where $t_j$ has type $\tau$.
%},
Additionally, $\M$ will interpret wrongly-applied selectors based on equality information in $\Fc$,
and will interpret the variables of $\Fc$ based on the valuation $\Val$ which was constructed in step two of our calculus.

Recall that $\Val$ is a mapping from the equivalence classes of $\Fc$ to $\mu$-terms.
We first \emph{complete} $\Val$ so that it maps equivalence classes only to closed $\mu$-terms, that is, terms containing no free variables.
This is done by exhaustively applying the following rule:

We write $\mathcal{T}(\Val)$ to denote the smallest set containing $\nf{(v\interp{u})}$ for each non-variable subterm $u$ of $v$ in the range of $\Val$.
%We write $\vrange(\Val)$ to denote the set of $\mu$-terms mapped to by $\Val$.
Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.

Let $\Val^\ast$ be the result of applying the above rule to completion on the mapping $\Val$ we constructed during step 2 of our calculus.
We first argue that $FV( \Val^\ast ) = \emptyset$.
Since \rn{Split} does not apply to $\Fc$,
all equivalence classes of $\Fc$ finite datatype type contain constructor terms, 
and thus $FV( \Val )$ contains no variables of finite datatype type.
For each equivalence classes of $\Fc$ with finite co-datatype $\tau$, 
we know $\tau$ must have cardinality one, due to \rem{reference}.
Since \rn{Singleton} does not apply to $\Fc$, 
there is only one equivalence class of type $\tau$, call it $\ec{x}$.
%It cannot be the case that $v_{\ec{x}} \in FV(\Val^\ast \ec{x})$, then we may apply \rn{Complete} for the (single) value $c$ of type $\tau$, 
%for which the second premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.
If it were the case that $v_{\ec{x}} \in FV(\Val^\ast )$,
then \rn{Complete} would apply for the (single) value $c$ of type $\tau$, 
for which the third premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.
Since all $\tau \in \Nondata$ are infinite, all variables in $FV( \Val^\ast )$ have infinite type.
For any infinite type $\tau$,
there exists a closed normal $\mu$-term $c_1$ that is not bisimilar to any instance of a term in $t \in \mathcal{T}(\Val^\ast)$.
To argue this is the case, notice that either $\tau \in \Nondata$,
$\tau$ is a recursive type, or contains an infinite subfield of type $\upsilon$.
In the first case, we may choose an arbitrary constant for $c_1$ not yet occurring in $\vrange(\Val)$.
In the second case, we choose a value $c_1$ whose depth is greater than any term in $\vrange(\Val)$.
In the third case, we choose a value $c_1$ containing a subterm of type $\upsilon$ that is not bisimilar to any instance of a term $\mathcal{T}(\Val^\ast)$, 
which is guarenteed to exist by applying this argument inductively.
%since $\mathcal{T}(\Val^\ast)$ is a finite set. \rem{elaborate} % (noting that a closed $\mu$-term is only bisimilar to other closed $\mu$-terms).
Thus, $FV( \Val^\ast )$ contains no variables of infinite type,
and consequently $\Val^\ast$ is a mapping from the equivalence classes of $\Fc$ to closed $\mu$-terms.
We further argue that $\Val^\ast$ has the following properties:
%the following invariants are maintained by applications of \rn{Complete}:% for all $\ec{x}, \ec{y}$:
\begin{eqnarray}
\label{eqn:m-normal} 
\Val^\ast \text{ maps to normal $\mu$-terms only } \\
%\text{If } \Val \ec{x} \text{ is not closed and has subterm $u$, then } \Val \ec{x} \interp{ u } = \Val \ec{y} \text{ for some } \ec{y} \\
%\Val \ec{x} = \Val \ec{y}\interp{u} \text{ for some $\ec{x}$, for all non-closed subterms } u \text{ of } \Val \ec{y},  \\
%\Val \ec{x} = v\interp{u} \text{ for some } \ec{x}, \text{ for all non-closed } u \in \mathcal{T}( \Val ) \\
%\Val \ec{x} \interp{\mu v_{\ec{y}}. t} = \Val \ec{y} \\
\label{eqn:m-bij} 
\Val^\ast \ec{x} \sim \Val^\ast \ec{y} \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}
We show this by induction on the number of applications of \rn{Complete}.
We additionally show that applications of \rn{Complete} maintain the invariant 
for all subterms $\mu v_{\ec{x}}. t$ occurring in the range of $\Val$ for some $\ec{y}$:
%\begin{eqnarray}
%\label{eqn:m-inv}
%\Val \ec{y} \interp{ \mu v_{\ec{x}}. t } = \Val \ec{x}
%\end{eqnarray}
For the base case, since \rn{Bisimilar} does not apply to $\Fc$,
we have that $\Val \ec{x} \not\vsim \Val \ec{y}$ for all distinct $\ec{x}$ and $\ec{y}$,
and thus~\ref{eqn:m-bij} holds for $\Val$.
\rem{todo 1}
%Due to our construction of $\Val$ in step 2 of our calculus, 
%statement~\ref{eqn:m-inv} holds, which together with~\ref{eqn:m-bij} implies~\ref{eqn:m-normal}.
%Moreover, since an equivalence class $\ec{u}$ exists for all subterms $u$ of all terms $t \in \mathcal{T}(\Fc)$,
%we have that $t\interp{u} = \Val \ec{u}$ for all such terms $u$ and $t$. \rem{elaborate?}
For the inductive case, we assume an arbitrary $\Val_i$ satisfies~\ref{eqn:m-normal} and~\ref{eqn:m-bij},
and show that $\Val_i \sigma_x$ does as well, where $\sigma_x := \{ v_{\ec{x}} \mapsto c_1 \}$ is a substitution applied by \rn{Complete}.
By contradiction, assume that $\Val_i \sigma_x \ec{y} \sim \Val_i \sigma_x \ec{z}$ for some distinct $\ec{y}$ and $\ec{z}$.
Then it must have been the case that $\Val_i \ec{y}$ and $\Val_i \ec{z}$ have a position $p$
where, w.l.o.g., say $\Val_i \ec{y} \mid_p$ is $v_{\ec{x}}$ and $\Val_i \ec{z} \mid_p$ is a term of the form $\mu v_{\ec{w}}. t$ 
and $c_1 \sim (\mu v_{\ec{w}}. t) \sigma_x$.
%Since $\Val_i$ satisfies invariant~\ref{eqn:m-bij},
%we have that $\Val_i \ec{w} \sim c_1$ for some $\ec{w}$, 
However, this contradicts the premise of \rn{Complete} that $c_1$ is not bisimilar to any $t \sigma \in \mathcal{T}( \Val_i )$ for any $\sigma$.
%Thus, by induction on the number of applications of \rn{Complete}, statements~\ref{eqn:m-normal} and~\ref{eqn:m-bij} hold for $\Val^\ast$.
Thus, statements~\ref{eqn:m-normal} and~\ref{eqn:m-bij} hold for $\Val^\ast$.

%we have that $\Val$ satisfies both~\ref{eqn:m-normal} and~\ref{eqn:m-bij}.
%It satisfies the former since if a $\mu$-term $\mu v_{\ec{x}}. t_1$ had a bisimilar subterm $\mu v_{\ec{y}}. t_2$,
%then the rule \rn{Bisimilar} would have applied to $x$ and $y$. \rem{elaborate?}
%For the inductive case, we assume $\Val$ satisfies~\ref{eqn:m-normal} and~\ref{eqn:m-bij},
%and show that $\Val \sigma$ does as well, where $\sigma := \{ v_{\ec{x}} \mapsto c_1 \}$ is a substitution applied by \rn{Complete}.
%By contradiction, assume that $\Val \sigma \ec{y}$ is a non-normal $\mu$-term for some $\ec{y}$.
%Since $c_1$ is normal, we have $\ec{x} \neq \ec{y}$, 
%and thus $\Val \ec{y}$ must have been of the form $\mu v_{\ec{y}}. t$ where $\ec{x} \in FV( t )$.
%Let $u$ be the (strict) subterm of $\Val \ec{y}$ such that $(\Val \ec{y} \sigma ) \interp{u \sigma } \sim (\Val \ec{y} \sigma)$. 
%Since $\Val$ maps only to normal $\mu$-terms, we have $x \in FV( u )$.

Now, let $\M( x )$ be the value in $\M$ that is a renaming of$\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$,
each of which are guarenteed to exist in $\M$ due to statement~\ref{eqn:m-normal} and our construction of $\M$,
and moreover are guarenteed to be distinct in $\M$ due to statement~\ref{eqn:m-bij}.
We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.

If $\tau \in \Nondata$, $t_1$ and $t_2$ must be variables, and thus $\M( t_1 ) = \M( t_2 )$ by construction.

\rem{revise for co-datatype}
If $\tau \in \Data \cup \Codata$,
we first show that $\M( t ) = \Val^\ast \ec{x}$ for all non-selector terms $t$ that occur in equivalence classes $\ec{x}$.
We show this by induction on the size of $\Val^\ast \ec{x}$.
In the case where $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t } = \Val^\ast \ec{ x }$.
In the case where $t$ is a constructor term $C( \vec u )$, 
then $\Val^\ast \ec{x}$ must have been assigned by \rn{Ind} for some constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By the induction hypothesis, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = C( \M( \vec u ) ) = C'( \M( \vec u ' ) ) = \Val^\ast \ec{ x }$.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}
\newcommand{\vsimv}[1]{\vsim_{#1}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node,
then $\Ec$ is satisfiable $\thD$.
To do so, we demonstrate a particular model $\M$ of $\thD$ that satisfies $\Ec$,
but first we require additional terminology concerning $\mu$-terms.

For a $\mu$-term $t_1$ with subterm $t_2$, 
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. \const{C}( \vec u ) \in t_1$ indicates that $\mu x. \const{C}( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
t_1\interp{x}^M & = & \mu x. \const{C}( t_1\interp{ \vec u}^{M \cup \{ x \}}) & x \not\in M, x \mapsto \mu x. \const{C}( \vec u ) \in t_1  \\
t_1\interp{x}^M & = & x & \text{otherwise} \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & \mu x. \const{C}( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & x \not\in M \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & x & \text{otherwise}
\end{array}\]
We will often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.
We will write $\mathcal{T}( t_1 )$ to denote the set consisting of $t_1\interp{t_2}$ for all subterms $t_2$ of $t_1$,
and write $\mathcal{T}( S )$ to denote the union of $\mathcal{T}( t )$ for each $t \in S$ where $S$ is a set of terms.

Given a $\mu$-term $t_1$ of the form $\mu x. \const{C}( u_1, \ldots, u_n )$ with subterm $t_2$,
we say $t_2$ is a \emph{bisimilar subterm} of $t_1$ if and only if it is of the form
$\mu y. \const{C}( v_1, \ldots, v_n )$ and $t_1\interp{ u_i } \sim t_1\interp{ v_i }$ for $i = 1,\ldots, n$.
We say $t_1$ is \emph{normal} if and only if it does not contain a strict bisimilar subterm
and all of its strict subterms are also normal.
For instance, the $\mu$-term $t_1$ of the form $\mu x. \const{C}( \mu y. \const{C}( y ) )$ is not normal, 
since $\mu y. \const{C}( y )$ is a bisimilar subterm of $t_1$.
In particular, note that their arguments have the same interpretation in $t_1$, 
that is,
$t_1\interp{\mu y. \const{C}( y )}^\emptyset =$ 
$\mu y. \const{C}( t_1\interp{y}^{\{y\}} ) =$ 
$\mu y. \const{C}( y )$ 
is syntactically equivalent to
$t_1\interp{y}^\emptyset =$ 
$\mu y. \const{C}( t_1\interp{y}^{\{y\}} ) =$ 
$\mu y. \const{C}( y )$.
The $\mu$-term $t_1$ of the form $\mu x. \const{C}( \mu y. \const{C}( x ) )$ is also not normal, 
since $\mu y. \const{C}( x )$ is a bisimilar subterm of $t_1$,
noting that
$t_1\interp{\mu y. \const{C}( x )}^\emptyset =$ 
$\mu y. \const{C}( t_1\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{C}( t_1\interp{\mu x. \const{C}( \mu y. \const{C}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{C}( \mu x. \const{C}( t_1\interp{\mu y. \const{C}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{C}( \mu x. \const{C}( y ) )$
is syntactically equivalent to $t_1\interp{x}^\emptyset = t_1$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form} of $v$)
that is a renaming of $v$ by replacing all of the bisimilar subterms of $v$ with $x$,
and by replacing all of the non-normal subterms of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{C}( \mu y. \const{C}( x ) ))} = \mu x. \const{C}( x )$.

We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing variables besides $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.

%\begin{lemma}
%If $t_1$ and $u$ are normal, and if $u \not\vsimv{x} t_1\interp{t_2}$ for all $t_2 \in \mathcal{T}( t_1 )$,
%then $t_1 \{ x \mapsto u \}$ is normal.
%\end{lemma}
%\begin{proof}
%\rem{TODO}
%\end{proof}

\begin{definition}[Normal Model] \label{def:norm-model}
A model $\M$ is \emph{normal} if and only if:
\begin{enumerate}
\item 
for all types $\tau$, 
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are 
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$,
\item 
for all constructor terms $\const{C}( \vec t )$ of type $\tau$,
$\M( \const{C}( \vec t ) )$ is the value 
in $\M( \tau )$ that is a renaming of 
$\nf{(\mu x. \const{C}( \M( \vec t ) ))}$ where $x$ is fresh, and
\item 
for all selector terms $s^j_i( t )$ of type $\tau$,% for some $\vec u$,
$\M( s^j_i( t ) )$ is the value
in $\M( \tau )$ that is a renaming of 
$\M( t )\interp{u_j}$
when $\M( t )$ is of the form $\mu x. \const{C}_i( \vec u )$.
\end{enumerate}
\end{definition}

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how wrong-applied selector terms and variables in $\Ec$ are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
Given a normal model $\M$, the \emph{completion} of $\Val$, which we will write as $\Val^\ast$, 
is the result of exhaustively applying the following update to $\Val$:

\(
\inferrule{
  x : \tau \in FV( \Val ) 
  \quad
  \mu x. t \sim v, v \in \M( \tau )
  \quad
  (\mu x. t) \not\sim_x u \text{ for all } u \in \mathcal{T}(\vrange(\Val))
}{
  \Val := \Val \{ x \mapsto \mu x. t \}
}
\)

Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
We first state the following properties of $\Val$, 
where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.
%which amounts to saying that 
%equivalence class are interpreted as values,
%the interpretation of an equivalence class is not sensitive to the context in which it is interpreted,
%and that each equivalence classes has a distinct interpretation.

\begin{lemma} 
\label{lem:model-completion}
The following are true of $\Val^\ast$:
\begin{enumerate}
\item $\nf{(\Val^\ast \ec{x})} \in \M( \tau )$ for all $\ec{x} : \tau$ of $\Ec$,
\item
$t_1 \interp{ t_2 } = \nf{(\Val^\ast \ec{x})}$
for all $t_1\interp{t_2} \in \mathcal{T}(\Val)$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\nf{(\Val^\ast \ec{x})} \vsim \nf{(\Val^\ast \ec{y})}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we must argue that $FV( \Val^\ast ) = \emptyset$, and that $\nf{(\Val^\ast \ec{x})}$ is not cyclic for
$\ec{x} : \tau$ when $\tau \not\in \Codata$.
\rem{TODO}
\qed
\end{proof}

@@

We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from step two of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is a renaming of $\nf{(\Val^\ast \ec{x})}$ for each variable $x$ in $\mathcal{T}(\Fc)$.
Notice that due to Lemma~\ref{lem:model-completion}.3, for variables $x, y \in FV( \Fc )$,
\begin{eqnarray}
\label{eqn:m-bij} 
\M( x ) = \M( y ) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show that $\M( t ) = \Val^\ast \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$, which we show by induction on the size of $t$.

If $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t }$ by construction.
If $t$ is a constructor term $\const{C}( u_1, \ldots, u_n )$, 
then since \rn{Unify$_2$} does not apply to $\Fc$,
then $\Val^\ast \ec{t}$ must be a term of the form $\mu v_{\ec{t}}. \const{C}( w_1, \ldots, w_n )$,
where say that $\muvar( w_i ) = v_{\ec{y_i}}$ for $i = 1, \ldots, n$.
Since \rn{Unify$_1$} does not apply to $\Fc$, it must be that $\ec{u_i} = \ec{y_i}$ for $i = 1, \ldots, n $.
%Due to Lemma~\ref{lem:model-completion}.2 and by induction
\rem{TODO}

%If $\tau \in \Nondata$, $t_1$ and $t_2$ must be variables, and thus $\M( t_1 ) = \M( t_2 )$ by construction.
%If $\tau \in \Data \cup \Codata$, we show $\M \models t_1 \teq t_2$ by showing 
%We show by induction on the term size of $t$ that $\M( t ) = \Val^\ast \ec{t}$.
%If $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t } = \Val^\ast \ec{ x }$.
%In the case where 
%where say $\muvar( v_i ) = v_{\ec{y_i}}$ for $i = 1, \ldots, n$.
%We have $\M( \const{C}( \vec u ) ) = \nf{(\mu y. \const{C}( \M( \vec u ) ))}$.
